Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-09 paper/galsize_paper.tex
Minor revisions; most blue-color text now removed and addressed

Using galaxies observed in the Sloan Digital Sky Survey (SDSS) with the stellar mass measurements taken from the MPA-JHU catalog \citep{kauffmann_etal03,brinchmann_etal04}, we define volume-limited samples of galaxies according to the same $\mstar-$completeness cuts used in \citet{behroozi_etal15} (see Figure 2). \textcolor{blue}{AK: which DR the catalog is taken from?} We supplement the MPA-JHU catalog with measurements of half-light radius, $\rhalf,$ derived from galaxy profile decompositions provided by \citet{meert_etal15}. The \citet{meert_etal15} catalog is based on Data Release 7 of SDSS \citep{abazajian_etal09}, with improvements to the photometry pipeline and light profile fitting method and, especially, in background subtraction \citep{vikram_etal10,bernardi_etal13,bernardi_etal14,meert_etal13}. In the version of this catalog that we use,\footnote{Our $\mstar$ and $\rhalf$ measurements are derived from different photometry pipelines; this is driven by our desire for consistency with the \citet{behroozi_etal15} completeness cuts used in our clustering measurements. We note that we have repeated our analysis for stellar mass measurements based on \citet{meert_etal15}, finding only minor quantitative, and no qualitative changes to our results.} two-dimensional $r-$band profiles were fit with a two-component de Vaucouleurs + exponential profile to determine the half-light radius of total $r-$band luminosity, $\rhalf.$

@@

The success of the $\rvir$-based model in explaining the size dependence of galaxy clustering implies that sizes of {\it all} galaxies,  red, blue, spheroid, disks, satellites and central, are, on average, proportional to the $\rmpeak,$ the physical size of the virial radius at the time at which $\mhalo$ stops growing. Although correlation is not necessarily causation, such universal and simple dependence poses a useful challenge to galaxy formation models, as it indicates tight connection betw
een growth of galaxies and their host halos.

The $\rvir$-based size model also assumes that after halo stops growing the size remains approximately constant and this assumption is consistent with observed size dependence of galaxy clustering in SDSS.  The halos of typical satellites reach $\mpeak$ far outside the virial radius of their ultimate host halo \citep{behroozi_etal14}, and so in the $\rvir-$based model the size difference between satellites and centrals is largely in place prior to the time of satellite infall. 
Thus, as noted above, halos of central and satellite galaxies of the same $\mstar$  differ in epochs when their halo stopped growing (i.e., $\zpeak$) and this difference explains their different median sizes.

The difference also explains the relative trends of $\rhalf-\mstar$ relations for red and blue, or disk and spheroid galaxies.
Both types of galaxies exhibit weak dependence of $\rhalf$ on $\mstar$ at $\mstar\lesssim 5\times 10^{10}\ M_\odot$ and 
at these masses $\rhalf$ of disks is a factor of two larger than those of spheroids. At larger masses, the sizes of galaxies of both types show stronger dependence on $\mstar$ and the difference in median size of disk and spheroid galaxies decreases \citep[e.g.,][]{bernardi_etal14}.

In the $\rvir$-based size model, the weak dependence of $\rhalf$ on $\mstar$ for low mass galaxies is due to steep $\mstar-\mvir$ relation. For example, if $\mstar\propto\mvir^\alpha\propto\rvir^{3\alpha}$, then $\rhalf\propto \rvir\propto \mstar^{1/(3\alpha)}$. Given that $\alpha\sim 1-3$ is inferred for the $\mstar-\mvir$ relation at low masses \citep{kravtsov10,behroozi13_smhm,kravtsov_etal14}, the $\rhalf-\mstar$ relation in this regime is expected to be shallow. 
For galaxies of $\mstar\gtrsim 5\times 10^{10}\ M_\odot$, $\alpha\approx 0.3$ \citep[e.g.,][]{kravtsov_etal14} and thus 
the slope of the $\rhalf-\mstar$ relation is approximately linear. These trends of $\rhalf$ with $\mstar$ have a direct relation to the trends of surface brightness $\mu_\star=\mstar/(\pi\rhalf^2)$ with stellar mass.

The difference in median sizes of disk and spheroidal galaxies at $\mstar\lesssim 10^{11}\ M_\odot$ is explained by the fact that sizes of satellite galaxies are systematically smaller than sizes of central galaxies of the same $\mstar$ and   
the satellite fraction is much higher among spheroidal galaxies compared to the disk galaxies. For example, analyses of \citet{rodriguez_puebla_etal15} show that in this regime satellite fraction of blue galaxies is only $\approx 10-20\%$, while
the satellite fraction of red galaxies is $\gtrsim 30-50\%$ increasing with decreasing stellar mass. 
As $\mstar$ increases, satellite fractions of both blue and red galaxies drop and at $\mstar\gtrsim 10^{11}\ M_\odot$ vast majority of blue and red galaxies are centrals. If these galaxies occupy similar halo masses at the same $\mstar$, $\rvir$-based size model predicts that  the difference in sizes of red and blue galaxies should disappear at these masses, as is indeed observed.

A testable prediction of this picture is that when 3D (de-projected) sizes of only central red and blue galaxies of the same $\mstar$ are compared they should be much closer, than the sizes of the overall red and blue populations.

@@

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/rvir_only_lensing_prediction.pdf}
\caption{
{\bf Prediction for $\rhalf-$dependence of galaxy lensing.}
Using the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}, we make predictions for as-yet-unseen measurements of the $\rhalf-$dependence of galaxy lensing of $\mstar-$complete samples, where ``small" and ``large" subsamples are defined as in \S\ref{subsec:sizedef}. As a generic consequence of satellites being smaller than centrals of the same mass, we predict that future lensing measurements of $\mstar-$complete samples will show $\Delta\Sigma$ of small galaxies to be significantly stronger relative to large galaxies of the same stellar mass. To date, the $\rhalf-$dependence of $\Delta\Sigma$ has only been measured for color-selected samples \citep{charlton_etal17}, in which the trend is reversed and weakened: for both blue and red samples, $\Delta\Sigma$ of large galaxies is (slightly) stronger relative to small galaxies. \textcolor{blue}{AK: is it worth showing a second panel with $\Delta\Sigma$ for color-selected samples to back up the last sentence?}
}
\label{fig:lensingprediction}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

On the one hand, this degeneracy with the satellite fraction is unfortunate, because it means galaxy size does not leave a pure and unique signature on $\rhalf-$clustering ratios. However, this can also be viewed as an opportunity to extract tighter constraints on $F_{\rm sat}(\mstar),$ which are sorely needed to discriminate between competing models \citep{watson_conroy13}. Traditional galaxy clustering is already being used to validate and/or fit models of the stellar-to-halo-mass relation \citep[e.g.,][]{leauthaud_etal11,moster_etal13,behroozi13_smhm,lehmann_etal15}. Based on our results, we advocate that empirical models for $\mean{\mstar}{\mhalo}$ be supplemented with additional model ingredients for $\median{\rhalf}{\mstar},$ (\textcolor{blue}{AK:} did you mean $\langle\rhalf|\mhalo\rangle$ here? Seems more logical in the context of this sentence) and that the parameters of the composite model be {\em jointly} constrained by measurements of the stellar mass function, $\mstar-$dependent clustering, and $\rhalf-$dependent clustering.

>>>>

Using galaxies observed in the Data Release 10 of the Sloan Digital Sky Survey \citep[SDSS,][]{ahn_etal14} with the stellar mass measurements taken from the MPA-JHU catalog \citep{kauffmann_etal03,brinchmann_etal04}, we define volume-limited samples of galaxies according to the same $\mstar-$completeness cuts used in \citet{behroozi_etal15} (see Figure 2). We supplement the MPA-JHU catalog with measurements of half-light radius, $\rhalf,$ derived from galaxy profile decompositions provided by \citet{meert_etal15}. The \citet{meert_etal15} catalog is based on Data Release 7 of SDSS \citep{abazajian_etal09}, with improvements to the photometry pipeline and light profile fitting method and, especially, in background subtraction \citep{vikram_etal10,bernardi_etal13,bernardi_etal14,meert_etal13}. In the version of this catalog that we use,\footnote{Our $\mstar$ and $\rhalf$ measurements are derived from different photometry pipelines; this is driven by our desire for consistency with the \citet{behroozi_etal15} completeness cuts used in our clustering measurements. We note that we have repeated our analysis for stellar mass measurements based on \citet{meert_etal15}, finding only minor quantitative, and no qualitative changes to our results.} two-dimensional $r-$band profiles were fit with a two-component de Vaucouleurs + exponential profile to determine the half-light radius of total $r-$band luminosity, $\rhalf.$

@@

The success of the $\rvir$-based model in explaining the size dependence of galaxy clustering implies that the sizes of {\it all} galaxies,  red, blue, spheroid, disks, satellites and central, are, on average, proportional to the $\rmpeak,$ the physical size of the virial radius at the time at which $\mhalo$ stops growing. Although correlation is not necessarily causation, such a universal and simple dependence poses a useful challenge to fine-grained galaxy formation models, as it indicates tight connection between the growth of galaxies and their host halos.

The $\rvir$-based size model assumes as an ansatz that galaxy size remains approximately constant after a halo stops growing; as shown in Figure \ref{fig:clustering_ratio_upshot}, this assumption is consistent with the observed size-dependence of galaxy clustering in SDSS.  The halos of typical satellites reach $\mpeak$ far outside the virial radius of their ultimate host halo \citep{behroozi_etal14}, and so in the $\rvir-$based model the size difference between satellites and centrals is largely in place prior to the time of satellite infall. Halos of central and satellite galaxies of the same $\mstar$ differ in the epoch $\zpeak$ at which their halo stopped growing, and this difference gives a natural explanation for their different median sizes.

This difference also explains the relative trends of $\rhalf-\mstar$ relations for red and blue, or disk and spheroid galaxies.
Both types of galaxies exhibit weak dependence of $\rhalf$ on $\mstar$ at $\mstar\lesssim 5\times 10^{10}\ M_\odot$ and
at these masses $\rhalf$ of disks is a factor of two larger than those of spheroids. At larger masses, the sizes of galaxies of both types show stronger dependence on $\mstar,$ and the difference in median size of disk and spheroid galaxies decreases \citep[e.g.,][]{bernardi_etal14}.

In the $\rvir$-based size model, the weak dependence of $\rhalf$ on $\mstar$ for low-mass galaxies is due to the steep $\mstar-\mvir$ relation. For example, if $\mstar\propto\mvir^\alpha\propto\rvir^{3\alpha}$, then $\rhalf\propto \rvir\propto \mstar^{1/(3\alpha)}$. Given that $\alpha\sim 1-3$ is inferred for the $\mstar-\mvir$ relation at low masses \citep{kravtsov10,moster_etal13,behroozi13_smhm,kravtsov_etal14}, the $\rhalf-\mstar$ relation in this regime is expected to be shallow.
For galaxies of $\mstar\gtrsim 5\times 10^{10}\ M_\odot$, $\alpha\approx 0.3$ \citep[e.g.,][]{kravtsov_etal14}, and thus
the slope of the $\rhalf-\mstar$ relation is approximately linear. These trends of $\rhalf$ with $\mstar$ have a direct relation to the trends of surface brightness $\mu_\star=\mstar/(\pi\rhalf^2)$ with stellar mass \citep[see, for example, the $\mu_\star-$dependent clustering measurements in][which are qualitatively similar to the measurements reported here]{li_etal06}.

The difference in median sizes of disk and spheroidal galaxies at $\mstar\lesssim 10^{11}\ M_\odot$ can be explained by the sizes of satellite galaxies being systematically smaller than the sizes of central galaxies of the same $\mstar,$ coupled with the fact that the satellite fraction is much larger among spheroidal galaxies compared to disk galaxies. For example, the analysis in \citet{rodriguez_puebla_etal15} shows that in this regime, the satellite fraction of blue galaxies is only $\sim10-20\%$, while
the satellite fraction of red galaxies is $\gtrsim 30-50\%,$ increasing with decreasing stellar mass.
As $\mstar$ increases, the satellite fractions of both blue and red galaxies drop, and at $\mstar\gtrsim 10^{11}\ M_\odot,$ the vast majority of blue and red galaxies are centrals. If these galaxies occupy similar halo masses at the same $\mstar$, then the $\rvir$-based size model predicts that the difference in sizes of red and blue galaxies should disappear at these masses, as is indeed observed. A testable prediction of this picture is that when 3D (de-projected) sizes of only central red and blue galaxies of the same $\mstar$ are compared, the sizes should be much closer than the sizes of the overall red and blue populations.

@@

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/rvir_only_lensing_prediction.pdf}
\caption{
{\bf Prediction for $\rhalf-$dependence of galaxy lensing.}
Using the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}, we make predictions for as-yet-unseen measurements of the $\rhalf-$dependence of galaxy lensing of $\mstar-$complete samples, where ``small" and ``large" subsamples are defined as in \S\ref{subsec:sizedef}. As a generic consequence of satellites being smaller than centrals of the same mass, we predict that future lensing measurements of $\mstar-$complete samples will show $\Delta\Sigma$ of small galaxies to be significantly stronger relative to large galaxies of the same stellar mass. To date, the $\rhalf-$dependence of $\Delta\Sigma$ has only been measured for color-selected samples \citep{charlton_etal17}, in which the trend is reversed and weakened: for both blue and red samples, $\Delta\Sigma$ of large galaxies is (slightly) stronger relative to small galaxies.
}
\label{fig:lensingprediction}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

On the one hand, this degeneracy with the satellite fraction is unfortunate, because it means galaxy size does not leave a pure and unique signature on $\rhalf-$clustering ratios. However, this can also be viewed as an opportunity to extract tighter constraints on $F_{\rm sat}(\mstar),$ which are sorely needed to discriminate between competing models \citep{watson_conroy13}. Traditional galaxy clustering is already being used to validate and/or fit models of the stellar-to-halo-mass relation \citep[e.g.,][]{leauthaud_etal11,moster_etal13,behroozi13_smhm,lehmann_etal15}. Based on our results, we advocate that empirical models for $\mean{\mstar}{\mhalo}$ be supplemented with additional model ingredients for $\median{\rhalf}{\mhalo},$ and that the parameters of the composite model be {\em jointly} constrained by measurements of the stellar mass function, $\mstar-$dependent clustering, and $\rhalf-$dependent clustering.
